---
title: 'Capstone Project: Case Study 1'
author: "Aleksandër Pulla"
date: "8/17/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Process

In order to complete this phase, the following tasks must be completed:

  1) Importing data into R 
  2) Combining datasets into a single data frame
  3) Ensuring consistency in columns
  4) Checking for errors
  5) Removing missing values
  6) Formatting data into appropriate datatypes
  7) Removing redundant data

#### Loading libraries and checking our working directory
```{r load libraries}
library(tidyverse)  # For wrangling and importing data
library(lubridate)  # For wrangling date functions
getwd()             # For displaying the current working directory

```

#### Importing datasets

```{r importing datasets}
april_2020 <- read.csv("2020/4 - April/2020_04_divvy_tripdata.csv") 
may_2020 <- read.csv("2020/5 - May/2020_05_divvy_tripdata.csv")
june_2020 <- read.csv("2020/6 - June/2020_06_divvy_tripdata.csv")
july_2020 <- read.csv("2020/7 - July/2020_07_divvy_tripdata.csv")
august_2020 <- read.csv("2020/8 - August/2020_08_divvy_tripdata.csv")
september_2020 <- read.csv("2020/9 - September/2020_09_divvy_tripdata.csv")
october_2020 <- read.csv("2020/10 - October/2020_10_divvy_tripdata.csv")
november_2020 <- read.csv("2020/11 - November/2020_11_divvy_tripdata.csv")
december_2020 <- read.csv("2020/12 - December/2020_12_divvy_tripdata.csv")
january_2021 <- read.csv("2021/1 - January/2021_01_divvy_tripdata.csv")
february_2021 <- read.csv("2021/2 - February/2021_02_divvy_tripdata.csv")
march_2021 <- read.csv("2021/3 - March/2021_03_divvy_tripdata.csv")
```

#### Exploring data

The script below displays several rows and all columns for each dataset

```{r displaying a brief overview of datasets, results='hide'}
head(april_2020)
head(may_2020)
head(june_2020)
head(july_2020)
head(august_2020)
head(september_2020)
head(october_2020)
head(november_2020)
head(december_2020)
head(january_2021)
head(february_2021)
head(march_2021)
```
The script below displays only columns names, in order to ensure their consistency between datasets columns

```{r diplay only column names to ensure their consistency}
colnames(april_2020)
colnames(may_2020)
colnames(june_2020)
colnames(july_2020)
colnames(august_2020)
colnames(september_2020)
colnames(october_2020)
colnames(november_2020)
colnames(december_2020)
colnames(january_2021)
colnames(february_2021)
colnames(march_2021)
```
The script below displays columns, their datatypes to ensure their consistency between datasets columns datatypes

```{r display columns, their datatypes to ensure consistency between datasets}
str(april_2020)
str(may_2020)
str(june_2020)
str(july_2020)
str(august_2020)
str(september_2020)
str(october_2020)
str(november_2020)
str(december_2020)
str(january_2021)
str(february_2021)
str(march_2021)
```

The script below will make sure that there are not extra white spaces in the observations

```{r removing whitespaces, results='hide'}
library(janitor)
clean_names(april_2020)
clean_names(may_2020)
clean_names(june_2020)
clean_names(july_2020)
clean_names(august_2020)
clean_names(september_2020)
clean_names(october_2020)
clean_names(november_2020)
clean_names(december_2020)
clean_names(january_2021)
clean_names(february_2021)
clean_names(march_2021)
```

Let's check the total of missing values for each dataset (from April 2020 - March 2021)

```{r total of NA values}
sum(is.na(april_2020))
sum(is.na(may_2020))
sum(is.na(june_2020))
sum(is.na(july_2020))
sum(is.na(august_2020))
sum(is.na(september_2020))
sum(is.na(october_2020))
sum(is.na(november_2020))
sum(is.na(december_2020))
sum(is.na(january_2021))
sum(is.na(february_2021))
sum(is.na(march_2021))
```

We can notice the existence of missing values for each dataset, especially for months like August, September, October (the month with the highest missing values) and November. 
It is time to remove this 'bad' data!

**Removing missing values (blanks, NA) from April 2020 dataset**
```{r Removing missing values (blanks, NA) from April 2020 dataset}
library(skimr)
april3 <- may_2020[!(is.na(april_2020$end_station_id) | april_2020$end_station_id == ""), ]
april_2020 <- april3
rm(april3)
# skim_without_charts(april_2020)
```

**Removing missing values (blanks, NA) from May 2020 dataset**

```{r Removing missing values (blanks, NA) from May 2020 dataset}
may3 <- may_2020[!(is.na(may_2020$end_station_id) | may_2020$end_station_id == ""), ]
may_2020 <- may3
rm(may3)
# skim_without_charts(may_2020)
```

**Removing missing values (blanks, NA) from June 2020 dataset**

```{r Removing missing values (blanks, NA) from June 2020 dataset}
june3 <- june_2020[!(is.na(june_2020$end_station_id) | june_2020$end_station_id == ""), ]
june_2020 <- june3
rm(june3)
# skim_without_charts(june_2020)
```

**Removing missing values (blanks, NA) from July 2020 dataset**

```{r Removing missing values (blanks, NA) from July 2020 dataset}
july2 <- july_2020[!(is.na(july_2020$start_station_id) | july_2020$start_station_id == ""), ]
july33 <- july2[!(is.na(july2$end_station_id) | july2$end_station_id == ""), ]
july_2020 <- july33
rm(july2)
rm(july33)
# skim_without_charts(july_2020)
```

**Removing missing values (blanks, NA) from August 2020 dataset**

```{r Removing missing values (blanks, NA) from August 2020 dataset}
august2 <- august_2020[!(is.na(august_2020$start_station_id) | august_2020$start_station_id == ""), ]
august3 <- august2[!(is.na(august2$end_station_id) | august2$end_station_id == ""), ]
august_2020 <- august3
rm(august2)
rm(august3)
# skim_without_charts(august_2020)
```

**Removing missing values (blanks, NA) from September 2020 dataset**

```{r Removing missing values (blanks, NA) from September 2020 dataset}
september2 <- september_2020[!(is.na(september_2020$start_station_id) | september_2020$start_station_id == ""), ]
september3 <- september2[!(is.na(september2$end_station_id) | september2$end_station_id == ""), ]
september_2020 <- september3
rm(september2)
rm(september3)
# skim_without_charts(september_2020)
```
**Removing missing values (blanks, NA) from October 2020 dataset**

```{r Removing missing values (blanks, NA) from October 2020 dataset}
october2 <- october_2020[!(is.na(october_2020$start_station_id) | october_2020$start_station_id == ""), ]
october3 <- october2[!(is.na(october2$end_station_id) | october2$end_station_id == ""), ]
october_2020 <- october3
rm(october2)
rm(october3)
# skim_without_charts(october_2020)
```

**Removing missing values (blanks, NA) from November 2020 dataset**

```{r Removing missing values (blanks, NA) from November 2020 dataset}
november2 <- november_2020[!(is.na(november_2020$start_station_id) | november_2020$start_station_id == ""), ]
november3 <- november2[!(is.na(november2$end_station_id) | november2$end_station_id == ""), ]
november_2020 <- november3
rm(november2)
rm(november3)
# skim_without_charts(november_2020)
```

**Removing missing values (blanks, NA) from December 2020 dataset**

```{r Removing missing values (blanks, NA) from December 2020 dataset}
december2 <- december_2020[!(is.na(december_2020$start_station_name) | december_2020$start_station_name == ""), ]
december3 <- december2[!(is.na(december2$start_station_name) | december2$end_station_name == ""), ]
december4 <- december3[!(is.na(december3$start_station_name) | december3$end_station_id == ""), ]
december_2020 <- december3
rm(december2)
rm(december3)
rm(december4)
# skim_without_charts(december_2020)
```

**Removing missing values (blanks, NA) from January 2021 dataset**

```{r Removing missing values (blanks, NA) from January 2021 dataset}
january1 <- january_2021[!(is.na(january_2021$start_station_name) | january_2021$start_station_name == ""), ]
january2 <- january1[!(is.na(january1$end_station_name) | january1$end_station_name == ""), ]
january_2021 <- january2
rm(january1)
rm(january2)
# skim_without_charts(january_2021)
```

**Removing missing values (blanks, NA) from February 2021 dataset**

```{r Removing missing values (blanks, NA) from February 2021 dataset}
february1 <- february_2021[!(is.na(february_2021$start_station_name) | february_2021$start_station_name == ""), ]
february2 <- february1[!(is.na(february1$end_station_name) | february1$end_station_name == ""), ]
february_2021 <- february2
rm(february1)
rm(february2)
# skim_without_charts(february_2021)
```

**Removing missing values (blanks, NA) from March 2021 dataset**

```{r Removing missing values (blanks, NA) from March 2021 dataset}
march1 <- march_2021[!(is.na(march_2021$start_station_name) | march_2021$start_station_name == ""), ]
march2 <- march1[!(is.na(march1$end_station_name) | march1$end_station_name == ""), ]
march_2021 <- march2
rm(march1)
rm(march2)
# skim_without_charts(march_2021)
```

Now that every dataset is cleaned, it is time to merge those into a single dataframe, called 'all_trips'. However, because some attributes have a particular datatype (i.e. 'chr') in some datasets and the same attributes but in other datasets have different datatype (i.e. 'int') it is important to convert every consistent attribute in the same datatype.

**Converting datatypes for each dataset**
```{r converting datatypes}
april_2020 <- mutate(april_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(april_2020)

may_2020 <- mutate(may_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(may_2020)

june_2020 <- mutate(june_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(june_2020)

july_2020 <- mutate(july_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(july_2020)

august_2020 <- mutate(august_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(august_2020)

september_2020 <- mutate(september_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(september_2020)

october_2020 <- mutate(october_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(october_2020)

november_2020 <- mutate(november_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
# skim_without_charts(november_2020)
```

Now that all datasets have consistent attributes it is time to merge them together into a single dataframe called 'all_trips'.

**Combine datasets into a single dataframe**
```{r combine datasets into a single file}
all_trips <- bind_rows(april_2020, may_2020, june_2020, july_2020, august_2020, september_2020, october_2020, november_2020, december_2020, january_2021, february_2021, march_2021)

all1 <- all_trips[!(is.na(all_trips$end_station_id) | all_trips$end_station_id == ""), ]
all_trips <- all1
rm(all1)
# skim_without_charts(all_trips)
```

Let's make sure there are not redundant data (duplicates).

**Checking for redundancy**
```{r checking for redundancy}
length(unique(all_trips$ride_id))
nrow(all_trips)
length(unique(all_trips$ride_id)) == nrow(all_trips) 
```
The result indicates that there are duplicate values.
Now it is time to find those kind of data and remove from the dataframe.

**Removing duplicates**
```{r removing duplicates}
all_trips <- distinct(all_trips)
#unique(all_trips)
#skim_without_charts(all_trips)
```
In the end of the 'PROCESS' phase let's create .csv files from these cleaned dataframes.

**Creating csv files from the cleaned dataframes**
```{r Creating csv files from the cleaned datasets}
write.csv(april_2020, "C:\\Users\\alexs\\Desktop\\april_2020.csv")
write.csv(may_2020, "C:\\Users\\alexs\\Desktop\\may_2020.csv")
write.csv(june_2020, "C:\\Users\\alexs\\Desktop\\june_2020.csv")
write.csv(july_2020, "C:\\Users\\alexs\\Desktop\\july_2020.csv")
write.csv(august_2020, "C:\\Users\\alexs\\Desktop\\august_2020.csv")
write.csv(september_2020, "C:\\Users\\alexs\\Desktop\\september_2020.csv")
write.csv(october_2020, "C:\\Users\\alexs\\Desktop\\october_2020.csv")
write.csv(november_2020, "C:\\Users\\alexs\\Desktop\\november_2020.csv")
write.csv(december_2020, "C:\\Users\\alexs\\Desktop\\december_2020.csv")
write.csv(january_2021, "C:\\Users\\alexs\\Desktop\\january_2021.csv")
write.csv(february_2021, "C:\\Users\\alexs\\Desktop\\february_2021.csv")
write.csv(march_2021, "C:\\Users\\alexs\\Desktop\\march_2021.csv")

write.csv(all_trips, "C:\\Users\\alexs\\Desktop\\all_trips.csv")
```


